---
layout: post
title: "PA Teacher Turnover"
output: github_document
---


## Background


T H I S   A   P R E L I M I N A R Y    D R A F T     

Teacher supply is a major issue facing the education system. This report focuses on teacher supply components in Pennsylvania.

## Methods


### Data
The Pennsylvania Department of Education (PDE) provides public data files that list teachers in public schools across the state. Because these teachers are provided with a unique identifier, and the data are available across multiple years, it is possible to examine teacher turnover of the Pennsylvania public teacher workforce. 

Data files are available from the PDE on their <a href="https://www.education.pa.gov/DataAndReporting/ProfSupPers/Pages/ProfPersIndStaff.aspx">Data and Reporting page</a>. Documentation for these data are available in PDE's <a href="https://www.education.pa.gov/Documents/Teachers-Administrators/PIMS/PIMS%20Manuals/2021-2022%20PIMS%20Manual%20Vol%201.pdf">PIMS Manual</a>.

In many cases, an individual teacher is included multiple times in the same year to reflect their assignments in different schools within a school district. Positions such as alternative education, computer science, art, ESL, and special education staff may teach in separate buildings in districts that need to allocate them as such. Only a teacher's 'primary assignment' record is included in the analysis file. [^2]

Individual teacher records for each of the years 2016-17, 2017-18, 2018-19, 2019-20 and 2020-21 were stacked into one data file holding nearly one million records. This file was filtered to include **teachers only**, thus, staff designated as 'Subcontracted,' 'Other Employee,' and 'Substitute' staff were not included in the analysis file. 

  
```{r, echo=F,message=FALSE,warning=FALSE,include=FALSE}
library(tidyverse);library(expss)
source("Z:/GENERAL/R/BC_conveniencefuncs.R", verbose = FALSE, print.eval = FALSE)
df<-read.csv("Z:/GENERAL/Data/PA_Data/PDE/PubSchoolStaff/paEducStaffRaw_SortedTchrsOnly_Merge.csv")

# clean Job Class

df$JobClass2<-
  ifelse(df$JobClass=="TPE","Temporary Professional Employee",
  ifelse(df$JobClass=="OE","Other Employee",
  ifelse(df$JobClass=="PE","Professional Employee",
  ifelse(df$JobClass=="SC","Subcontracted Employee",
  ifelse(df$JobClass=="",NA,
  df$JobClass)))))

df$YearsInLEA_cat<-ifelse(df$YearsInLEA==1,"One Year in LEA",
                          ifelse(df$YearsInLEA<5,"2-5 Years",
                                 ifelse(df$YearsInLEA<10,"5-10 Years",
                                        "10+ Years")))
df$YearsInLEA_cat<-factor(df$YearsInLEA_cat,
                          levels=c("One Year in LEA", "2-5 Years","5-10 Years","10+ Years"),ordered = T)

```

## Data Preparation
The code below does the following:

1. Read each year of PDE staffing data and format variables consistently in each year
2. Stack each year of data into one file
3. Save the file as CSV for analysis


The `paEducStaffRaw` file was also merged with district characteristics data from the National Center for Education Statistics using AUN.


**Read data files** (two examples provided)
```{r readFiles,eval=FALSE}
t16 <- read.csv("//filesvr01/yourpath/PDE/PubSchoolStaff/2016-17 Professional Personnel Individual Staff Report.csv")
colnames(t16)<-str_replace_all(colnames(t16)," ","")
colnames(t16)<-str_replace_all(colnames(t16),"[:punct:]","")
t16$year<-'16_17'
t16$PublicID<-as.character(t16$PublicID)
t16$PositionRecode<-ifelse(str_detect(t16$Position,"Teacher"),"Teacher",
                           ifelse(str_detect(t16$Position,"Administ"),"Admin",
                           t16$Position))
t16$YearsInED<-t16$YearsInEd
t16$PublicID<-as.numeric(t16$PublicID)  # subsequent years are numeric format
t16$AUN<-as.character(t16$AUN)
t16 %>% dplyr::filter(FirstName=="KAREN" & LastName=="KUNSA") %>%
  select(LastName,PublicID)

t17 <- read_excel("Z:/yourpath/2017-18 Professional Personnel Individual Staff Report.xlsx")
colnames(t17)<-str_replace_all(colnames(t17)," ","")
colnames(t17)<-str_replace_all(colnames(t17),"-","")
t17$year<-'17_18'
myvars<-c("AnnualSalary","YearsInEd","YearsInLEA")
t17[myvars]<-makenum(t17,myvars) # convert to numeric format
t17$PositionDescription<-t17$PositionDescription2017
t17$PositionRecode<-ifelse(str_detect(t17$PositionDescription,"Teacher"),"Teacher",
                           ifelse(str_detect(t17$PositionDescription,"Administ"),"Admin",
                                  t17$PositionDescription))
t17$YearsInED<-t17$YearsInEd
t17$AUN<-as.character(t17$AUN)

```

\ 

**Stack annual files** A subset of variables is included in the stacked, multiyear file.

```{r, eval=FALSE}
myvars<-c(
  "year", "FirstName","LastName","PublicID","YearsInED",
  "YearsInLEA","AUN","School",
  "PositionRecode","PrimaryAssignment","Status",  # not on earlier files
  "LEACountyCd", "JobClass"     )

t16a<-t16[myvars];
t17a<-t17[myvars];
t18a<-t18[myvars];t19a<-t19[myvars];t20a<-t20[myvars]

# bind most recent years of data
df<-rbind(t16a,t17a,t18a,t19a,t20a);  rm(t13a,t14a,t15a,t16a,t17a,t18a,t19a,t20a)

df<-df %>% arrange(PublicID,year)  # sort order is very important

```

\ 


T H I S   A   P R E L I M I N A R Y    D R A F T     


### Analytic Variables
The analysis uses variables that classify teachers' movement or attrition from an LEA. I used definitions from the National Center for Education Statistics to create key analytic variables for whether a teacher was a "leaver," a "mover," or a "stayer." Specifically, from the <a href="https://ies.ed.gov/ncee/edlabs/regions/northeast/pdf/REL_2021080.pdf">,report</a> *Analyzing Teacher  Mobility and Retention: Guidance and Considerations Report 1* these definitions are:

+ Leavers. Teachers who left their initial LEA between any two years and did not enter another
LEA during this timeframe.
+ Movers. Teachers who moved from their initial LEA to a different LEA in any two year period.	The LEA (the LEA a teacher is employed within) that determines whether a teacher is a mover might differ from the one used to determine whether a teacher is a leaver .	For example, a mover might be defined as a teacher who moves between LEAs, whereas a leaver might be defined as a teacher who is no longer teaching in the state. Note that a shortcoming of this analysis is that teachers who move to another state are not accounted for.
+ Stayers. Teachers who remained teaching in their initial LEA in any two year period.	*Further work needed on this variable!*

```{r}
df$StayMove_or_NA<-
  ifelse(df$PublicID==data.table::shift(df$PublicID,n=1,type='lead') & #next record must be same teacher
         df$AUN==data.table::shift(df$AUN,n=1,type='lead')  , 'Stayed in LEA',
         
  ifelse(df$PublicID==data.table::shift(df$PublicID,n=1,type='lead') &
         df$AUN!=data.table::shift(df$AUN,n=1,type='lead')  , 'Moved LEA',
  
  ifelse(df$PublicID!=data.table::shift(df$PublicID,n=1,type='lead'),"NA: Next Record Diff Tchr",
         NA)))

df$leaver<-
  ifelse(df$year=="2016_17" & df$sequence<5,"Left within 4 yrs",
         ifelse(df$year=="2017_18" & df$sequence<4,"Left within 3 yrs",     
         ifelse(df$year=="2018_19" & df$sequence<3,"Left within 2 yrs",
                ifelse(df$year=="2019_20" & df$sequence<2,"Left within 1 yrs","Other"))))

df=apply_labels(df,
                year="School Year",
                StayMove_or_NA="Teacher Status in Subsequent Year",
                charter="Charter School Status",hipov="High Poverty Status",
                 sameLEA2="Employment Status", YearsInLEA_cat="Yrs in LEA",
                changedLEA="Staff Changed LEA (look behind)", noReturn_paLEA="Staff Did Not Return (look ahead)")


```

\ 



## Analysis
This preliminary analysis begins with a simple question: *How many teachers remain in the staffing file from one year to the next?*

The answer to this question is best addressed by first examining the number of teachers in each year of the dataset.


```{r,echo=FALSE,message=FALSE,warning=FALSE}

fre(df$year) %>% 
  set_caption(tabtitle("Number of Teachers by Year")) %>% 
   htmlTable(., css.cell = c("width: 250px", # first column width
                              rep("width: 50px", ncol(.) - 1)) # other columns width
    )
```

\
 
We can also see the mobility status for all the teachers in the datafile:

```{r,echo=FALSE,message=FALSE,warning=FALSE}
df$StayMove_or_NA<-factor(df$StayMove_or_NA, levels=c("Stayed in LEA","Moved LEA","NA: Next Record Diff Tchr"),ordered = T)

cro(df$year, df$StayMove_or_NA ) %>% 
  set_caption(tabtitle("Teachers' Mobility Status by Year")) %>% 
   htmlTable(., css.cell = c("width: 250px", # first column width
                              rep("width: 50px", ncol(.) - 1)) # other columns width
    )
```

\ 

Next, I  queried whether a set of unique teacher IDs in one year appeared in a subsequent year. Specifically, the analysis below creates a set of `PublicID` values for each year and counts the number of teachers in that set to the values in a subsequent year using the following code:

`ID1617<-df %>% dplyr::filter(year=="16_17") %>% pull(PublicID) # create set of IDs for 1617`

`y2<-df %>% dplyr::filter(year=="17_18") # create the set of cases for 1718 `

`table(ID1617 %in% y2$PublicID) # how many IDs from 1617 in 2017-18?`


```{r TrackUniqueIDs, echo=F}
# for each year, how many IDs disappear by the following year?
ID1617<-df %>% dplyr::filter(year=="16_17") %>% pull(PublicID) # create set of IDs for 1617
ID1718<-df %>% dplyr::filter(year=="17_18") %>% pull(PublicID) # create set of IDs for 1718
ID1819<-df %>% dplyr::filter(year=="18_19") %>% pull(PublicID)
ID1920<-df %>% dplyr::filter(year=="19_20") %>% pull(PublicID)

y1<-df %>% dplyr::filter(year=="16_17") # Cases for 1617
y2<-df %>% dplyr::filter(year=="17_18") # Cases for 1718 
y3<-df %>% dplyr::filter(year=="18_19") 
y4<-df %>% dplyr::filter(year=="19_20") 
y5<-df %>% dplyr::filter(year=="20_21") 

# for each year, how many IDs are new?
#table(ID1617 %in% y2$PublicID) # how many IDs from 1617 in 2017-18?
#table(ID1617 %in% y3$PublicID) # how many IDs from 1617 in 2018-19?
#table(ID1617 %in% y4$PublicID) # how many IDs from 1617 in 2019-20?

pct1617<-table(ID1617 %in% y2$PublicID) # how many IDs from 1617 in 2017-18?
pct1617<-(pct1617[1]/pct1617[2])*100 # pct in 1617 
pct1718<-table(ID1718 %in% y3$PublicID) # how many IDs from 1718 appear in 2018-19?
pct1718<-(pct1718[1]/pct1718[2])*100 # pct in 1718 
pct1819<-table(ID1819 %in% y4$PublicID) # how many IDs from 1819 in 2019-20?
pct1819<-(pct1819[1]/pct1819[2])*100 # pct in 1819 
pct1920<-table(ID1920 %in% y5$PublicID) # how many IDs from 1920 in 2020-21?
pct1920<-(pct1920[1]/pct1920[2])*100 # pct in 1718 

t<-rbind(pct1617,pct1718,pct1819,pct1920)

```

\ 

For example, in 2016-17, there were `r df %>% dplyr::filter(year=="16_17") %>% count()` teachers in the analysis file and about `r round(pct1617,2)` percent "left" teaching in Pennsylvania by 2017-18.

------------------------------
   Year       % Not Continuing 
------------- ----------------
 **2016-17**       `r round(pct1617,2)` 

 **2017-18**      `r round(pct1718,2)`

 **2018-19**       `r round(pct1819,2)`

 **2019-20**       `r round(pct1920,2)`
------------------------------

Note that the results just above use a different method to calculate the percentages and cannot be directly compared to the stayer/mover figures.



[^1]: There are many news items describing teacher supply issues. For example, see <a href="https://www.readingeagle.com/2021/12/27/substitute-teacher-shortage/">McKeel (2021)</a> and <a href="https://www.nbcnews.com/politics/joe-biden/biden-wants-fix-nation-s-teacher-shortage-educators-say-problem-n1269340">Moore (2021)</a>.

[^2]: Often these 'other' assignments are for langauge, arts, technology and physical education teachers.
